\documentclass{article}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{textcomp}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{longtable}

\lstset{
    language=Java,
    basicstyle=\ttfamily\small,
    keywordstyle=\color{blue},
    stringstyle=\color{red},
    commentstyle=\color{green!60!black},
    breaklines=true,
    showstringspaces=false
}

\title{COS 710: Artificial Intelligence \\[1ex]
\large Assignment 2: Structure-Based Genetic Programming for Electricity Load Prediction}

\author{Aryan Mohanlall \\ U23565536}
\date{}

\begin{document}

\maketitle

\section{Description of the Data Set}

The data set used in this assignment is the \textit{Residential Energy Dataset UK 2014--2020}. It contains residential energy and weather measurements sampled at 15-minute intervals. The data set contains 201,604 rows. Each row corresponds to one timestamp and contains the following columns:

\begin{itemize}
    \item \texttt{utc\_timestamp}: timestamp of the reading.
    \item \texttt{Electricity\_load}: residential electricity load. This is the prediction target.
    \item \texttt{Residential\_electricity\_price}: residential electricity price.
    \item \texttt{Residential\_solar\_generation}: residential solar generation.
    \item \texttt{Residential\_wind\_generation}: residential wind generation.
    \item \texttt{Temperature}: ambient temperature.
    \item \texttt{Relative Humidity}: relative humidity.
\end{itemize}

The problem is treated as a time-series prediction task. The aim is to predict the current electricity load at time $t$, denoted $load_t$, without giving the predictor access to the current value of \texttt{Electricity\_load}. Therefore, \texttt{Electricity\_load} is used only as the target value, not as an input feature for the same row.

Since the data is sampled every 15 minutes, there are 96 readings per day:

\begin{equation}
24 \times 4 = 96.
\end{equation}

The input features are constructed from the current non-target variables and from the same-time records of the previous seven days, so $n=7$ in the assignment wording. The first seven days are therefore skipped because they do not have a complete history. The feature set used by each program is:

\begin{itemize}
    \item Current time features:
    \begin{center}
    \texttt{price\_t, solar\_t, wind\_t, temp\_t, humidity\_t, day, month, year, hour, minute}
    \end{center}

    \item Same-time features from the previous seven days:
    \begin{center}
    \texttt{load\_d1, price\_d1, solar\_d1, wind\_d1, temp\_d1, humidity\_d1}
    \end{center}
    through
    \begin{center}
    \texttt{load\_d7, price\_d7, solar\_d7, wind\_d7, temp\_d7, humidity\_d7}.
    \end{center}
\end{itemize}

This gives 52 input variables in total. The data is split chronologically, with the first 80\% used for training and the remaining 20\% used for testing. The data is not shuffled, because shuffling would break the time-series ordering and could leak future information into training.

\section{Description of the Performance Metric}

The quality of the predictor is assessed using Root Mean Squared Error (RMSE). For $N$ examples, with actual load $y_i$ and predicted load $\hat{y}_i$, RMSE is defined as:

\begin{equation}
\text{RMSE} = \sqrt{\frac{1}{N}\sum_{i=1}^{N}(y_i - \hat{y}_i)^2}.
\end{equation}

RMSE was selected because the task is a regression problem and the prediction error should be measured in the same unit as the target variable. RMSE also penalises large errors more strongly than Mean Absolute Error due to the squared term. This is appropriate for load forecasting, where large prediction errors are more undesirable than small deviations.

During evolution, the selection fitness includes a bloat penalty:

\begin{equation}
fitness(p) = RMSE(p) + \lambda \cdot size(p),
\end{equation}

where $size(p)$ is the number of nodes in program $p$ and $\lambda = 0.00001$. This penalty discourages unnecessarily large trees. The final reported training and test performance are still reported as pure RMSE values so that predictive accuracy can be interpreted directly.

\section{Description of How Structure is Incorporated into Canonical GP}

Canonical GP uses fitness as the main guide for search. In contrast, Structure-Based Genetic Programming (SBGP) uses both behavioural information, in the form of fitness, and structural information, in the form of the syntax tree shape.

The SBGP implementation follows the global--local structure described in the course material. Evolution alternates between two phases:

\subsection{Global Level Search}

The global phase performs canonical GP-style evolution for $S_g = 10$ generations. At the end of the global phase, the best individual is used to define a \textit{global area}. This global area is the top $D_g = 4$ levels of the best program tree.

For the second and later global phases, offspring are compared against previously exploited global areas. If the top $D_g$ levels of an offspring are too similar to a previously exploited global area, the offspring is discarded. Similarity is measured by counting matching nodes at the same positions in the two top-level tree structures. The threshold used is:

\begin{equation}
T_g = 6.
\end{equation}

This prevents the search from repeatedly returning to the same structural region.

\subsection{Local Level Search}

After a global area is selected, local search is performed within that structural region. The first $D_g$ levels of every individual are fixed to match the selected global area. Genetic operators are then only allowed to modify nodes below the fixed global area. This means that local search exploits a particular structural region while still allowing lower-level subtrees to evolve.

Local search terminates early if no improvement is observed for $W_g = 10$ generations. The algorithm then returns to global search to seek a different structural region.

\section{Description of the Representation Used and the Terminal and Function Sets}

\subsection{Program Representation}

Each individual is represented as a binary syntax tree. Internal nodes are function nodes and leaf nodes are terminals. A program is evaluated recursively from the leaves to the root to produce a numeric prediction for the electricity load.

The implementation uses the following node classes:

\begin{itemize}
    \item \texttt{Function}: internal node containing an operator and two child nodes.
    \item \texttt{Variable}: terminal node that reads one input feature.
    \item \texttt{Terminal}: terminal node containing an ephemeral random constant.
\end{itemize}

The root of every tree must be a function node, and each tree must contain at least three nodes. Therefore, a single variable such as \texttt{load\_d7} is not a valid program by itself. The smallest valid program has the form:

\begin{center}
\texttt{(x + y)}
\end{center}

where $x$ and $y$ may be variables or constants.

\subsection{Terminal Set}

The terminal set consists of:

\begin{itemize}
    \item the 52 input variables described in the data set section, and
    \item ephemeral random constants sampled uniformly from $[-1, 1]$.
\end{itemize}

The terminal set was chosen to give the GP access to variables that are available at prediction time and to avoid leakage from the value being predicted. The current non-load weather and generation variables were included because they describe the present operating conditions that may influence electricity demand. Calendar variables were included because residential load is time dependent: demand can differ by month, day, hour, and minute. Lagged values from the previous seven days were included because residential demand has strong daily and weekly patterns, so the same time on previous days is a useful predictor.

The current electricity load value is not included in the terminal set, because it is the target value to be predicted. Including \texttt{load\_t} as an input would make the task trivial and would leak the answer into the program. Ephemeral constants are included so that programs can learn offsets, thresholds, and scaling factors rather than only combining raw variables.

\subsection{Function Set}

The function set is:

\begin{equation}
\mathcal{F} = \{+, -, \times, \div, min, <0\}.
\end{equation}

All functions are implemented as binary operators. Division is protected: if the denominator is close to zero, the result returned is 1. The \texttt{min} operator returns the smaller of its two arguments. The \texttt{<0} operator is implemented as a binary conditional:

\begin{equation}
<0(a,b) =
\begin{cases}
b, & a < 0,\\
0, & otherwise.
\end{cases}
\end{equation}

\section{Description of Initial Population Generation}

The initial population is generated using ramped half-and-half. This method combines the full and grow tree-generation methods over a range of depths. The aim is to create an initial population with a mixture of small and larger trees, as well as a mixture of regular and irregular tree shapes.

\subsection{Full Method}

In the full method, all internal nodes up to the selected depth are function nodes, and terminals are only placed at the final level of the tree. Since all functions in this implementation are binary, the full method creates balanced binary trees. This encourages the initial population to contain larger and more structurally complete programs.

\subsection{Grow Method}

In the grow method, each non-root node may become either a function node or a terminal node, subject to the maximum depth limit. This allows trees to stop growing before the maximum depth is reached and therefore creates more irregular program shapes. The root is still forced to be a function node so that a single terminal such as \texttt{load\_d7} cannot be a valid complete program.

\subsection{Ramped Half-and-Half}

For ramped half-and-half, the selected depth cycles through the configured range from the minimum depth to the maximum depth. The implementation alternates between full and grow generation. For example, with depth range 2--6, individuals are generated across depths 2, 3, 4, 5, and 6, with alternating full and grow trees. This provides diversity in both depth and shape.

Duplicate trees are not permitted in the initial population. Each generated tree is converted to its string representation and compared against the trees already generated. If a duplicate is found, a new tree is generated using the same depth and generation method.

\section{Description of the Fitness Function and Fitness Evaluation}

Each program is evaluated on the training set. For every training example, the program tree receives the 52 input variables and returns a predicted electricity load. The prediction is compared to the actual load using squared error. The RMSE is then calculated over all training examples.

For evolutionary selection, the fitness function includes a bloat penalty:

\begin{equation}
fitness(p) = RMSE(p) + 0.00001 \cdot size(p).
\end{equation}

Lower fitness is better. If a program produces a non-finite value, such as infinity or NaN, it is assigned a very large error. This prevents invalid mathematical expressions from being selected.

The bloat penalty is used only to guide evolution. Final reported training and test scores are pure RMSE values.

\section{Description of the Selection Method}

Tournament selection is used. A tournament consists of four randomly selected individuals:

\begin{equation}
k = 4.
\end{equation}

The individual with the lowest fitness is selected. If two individuals have equal fitness, the individual with fewer nodes is preferred. This tie-breaking rule encourages smaller programs and further reduces bloat.

The tournament size is fixed at 4 rather than being scaled with the population size. This was chosen to keep the selection pressure moderate across all tested population sizes. Very large tournaments can make high-fitness individuals dominate too quickly and can increase the risk of premature convergence, especially in GP where maintaining structural diversity is important.

The implementation uses steady-state replacement. In each generation, one offspring is created and evaluated. If the offspring is better than the worst individual in the population, it replaces that worst individual. The population is sorted after replacement so that the best individual is kept at index 0.

Elitism is also used. The best program found so far is tracked and reinserted into the population so that it cannot be lost due to crossover, mutation, or hoist.

\section{Description of Genetic Operators}

The genetic operators used are reproduction, crossover, mutation, and hoist. The baseline rates used in the experiments are:

\begin{itemize}
    \item Crossover rate: 0.70
    \item Mutation rate: 0.15
    \item Hoist rate: 0.02
    \item Reproduction rate: $1 - 0.70 - 0.15 - 0.02 = 0.13$
\end{itemize}

\subsection{Reproduction}

Reproduction copies a selected parent unchanged. It preserves good programs and provides stability during evolution.

\subsection{Depth-Aware Crossover}

Crossover replaces a subtree in one parent with a subtree from another parent. Unlike standard subtree crossover, the implementation is depth-aware. After selecting the replacement point in the first parent, the algorithm calculates how much depth is still available before reaching the maximum allowed tree depth. The donor subtree from the second parent must fit within this remaining depth budget.

Crossover is also structure-aware. A function position is replaced by a function subtree, and a terminal position is replaced by a terminal subtree. This preserves the structural role of nodes and prevents invalid offspring. During local search, crossover is only allowed to edit nodes below the fixed global area.

\subsection{Structure-Aware Mutation}

Mutation depends on the type of node selected:

\begin{itemize}
    \item If a function node is selected, either the operator is changed while keeping its children, or the function subtree is replaced with a newly grown function-root subtree.
    \item If a terminal node is selected, it is replaced with another terminal, either a variable or an ephemeral random constant.
\end{itemize}

The mutation is restricted by the maximum tree depth. During local search, mutation is only applied below the fixed global area.

\subsection{Hoist}

Hoist selects a subtree and promotes one of its inner subtrees to replace it. This operator reduces tree size and combats bloat. The implementation rejects hoist results that would violate the minimum size, functional root, or maximum depth constraints.

\subsection{Similarity-Based Rejection}

During global search, offspring are compared to previously exploited global areas. The similarity score counts nodes that match at the same positions within the first $D_g$ levels. If the similarity score is greater than or equal to $T_g = 6$, the offspring is discarded. This is the main mechanism by which SBGP avoids revisiting previously explored structural regions.

\section{Experimental Setup}

The experiments were designed to evaluate the performance of Structure-Based Genetic Programming on the electricity load prediction task. The implementation was written in Java and each run trained an independent SBGP population on the same training partition of the data set. The final model from each run was then evaluated on the held-out test partition.

\subsection{Parameter Values}

The baseline parameter configuration used in the experiments is shown in Table~\ref{tab:parameters}. The values are consistent with the implementation in \texttt{Main.java} and \texttt{SBGP.java}.

\begin{table}[h]
\centering
\begin{tabular}{|l|l|}
\hline
\textbf{Parameter} & \textbf{Value} \\
\hline
Number of runs & 15 \\
Population size & 100 \\
Generations & 100 \\
Initial population method & Ramped half-and-half \\
Minimum initial depth & 2 \\
Maximum initial depth & 6 \\
Crossover rate & 0.70 \\
Mutation rate & 0.15 \\
Hoist rate & 0.02 \\
Reproduction rate & 0.13 \\
Tournament size & 4 \\
Bloat penalty & 0.00001 \\
Training split & 80\% \\
Testing split & 20\% \\
\hline
\end{tabular}
\caption{Baseline experimental parameter values.}
\label{tab:parameters}
\end{table}

The SBGP-specific parameters are shown in Table~\ref{tab:sbgp-parameters}. These values control the alternation between global exploration and local exploitation.

\begin{table}[h]
\centering
\begin{tabular}{|l|l|l|}
\hline
\textbf{Parameter} & \textbf{Symbol} & \textbf{Value} \\
\hline
Global search generations & $S_g$ & 10 \\
Global area depth & $D_g$ & 4 \\
Global similarity threshold & $T_g$ & 6 \\
Local window tolerance & $W_g$ & 10 \\
\hline
\end{tabular}
\caption{Structure-based GP control parameters.}
\label{tab:sbgp-parameters}
\end{table}

Fitness was evaluated using the full training set rather than a random sample. This was done to avoid sampling noise and to make the training fitness deterministic for a given program. If the full evaluation becomes too computationally expensive for larger parameter sweeps, the implementation can be adapted to use sampled training evaluation, but the results reported for this configuration are based on full training evaluation.

\subsection{Technical Specifications}

All experiments were implemented in Java (JDK 25) and executed on a single machine in a single threaded manner with the following specifications:

\begin{table}[h]
\centering
\begin{tabular}{ll}
\hline
\textbf{Component} & \textbf{Specification} \\
\hline
Processor           & Intel Core i7-13700H (13th Gen) \\
Base clock speed    & 2.40 GHz \\
Cores / Threads     & 14 cores / 20 logical processors \\
RAM                 & 32GB \\
Operating system    & Windows 11 \\
Language / Runtime  & Java JDK 25 \\
\hline
\end{tabular}
\caption{Technical specifications of the machine used for development and simulation.}
\label{tab:specs}
\end{table}

\section{Results and Discussion}

The SBGP experiments were run for 15 independent runs per configuration. This satisfies the minimum requirement of 10 runs and matches the experimental structure used for the canonical GP experiments in Assignment 1. The same data set, train/test split, population settings, depth settings, and operator-rate sweep were used so that the results are directly comparable. Lower RMSE values are better.

\begin{table}[h]
\centering
\resizebox{\textwidth}{!}{
\begin{tabular}{llccccc}
\hline
\textbf{\#} & \textbf{Configuration} & \textbf{Mean Train} & \textbf{Mean Test} & \textbf{StdDev Train/Test} & \textbf{Best Train} & \textbf{Best Test} \\
\hline
1 & Baseline & 0.02952 & 0.03513 & 0.01082 / 0.00986 & 0.01927 & 0.02450 \\
2 & pop=50 & 0.04488 & 0.05008 & 0.01956 / 0.02089 & 0.02014 & 0.02720 \\
3 & pop=200 & 0.02464 & 0.03089 & 0.00793 / 0.00647 & 0.01927 & 0.02658 \\
4 & depth=2-4 & 0.03120 & 0.03798 & 0.01149 / 0.01048 & 0.02096 & 0.02796 \\
5 & depth=2-8 & 0.02986 & 0.03450 & 0.01169 / 0.00831 & 0.02050 & 0.02753 \\
6 & cx=0.80, mut=0.10 & 0.03200 & 0.03787 & 0.01337 / 0.01509 & 0.01927 & 0.02484 \\
7 & cx=0.50, mut=0.30, hoist=0.10 & 0.02695 & 0.03278 & 0.01003 / 0.00710 & 0.01927 & 0.02658 \\
\hline
\end{tabular}}
\caption{SBGP RMSE results across 15 runs per configuration. Baseline = (pop=100, cx=0.70, mut=0.15, hoist=0.02, depth=2-6).}
\label{tab:sbgp-results}
\end{table}

The best mean test performance for SBGP was obtained by the larger population configuration, \texttt{pop=200}, with a mean test RMSE of 0.03089. This suggests that the larger population helped SBGP explore more structural regions. The best individual test result was obtained by the baseline configuration, with a best test RMSE of 0.02450. The \texttt{cx=0.80, mut=0.10} configuration also found a strong individual solution with a best test RMSE of 0.02484. These best values show that SBGP was able to find competitive individual solutions, but the mean and standard deviation values show that it did not do so reliably for every run.

\begin{table}[h]
\centering
\resizebox{\textwidth}{!}{
\begin{tabular}{llcccc}
\hline
\textbf{\#} & \textbf{Configuration} & \textbf{A1 Mean Test} & \textbf{A2 SBGP Mean Test} & \textbf{Difference} & \textbf{Better Mean} \\
\hline
1 & Baseline & 0.02729 & 0.03513 & +0.00784 & A1 \\
2 & pop=50 & 0.02912 & 0.05008 & +0.02096 & A1 \\
3 & pop=200 & 0.02674 & 0.03089 & +0.00415 & A1 \\
4 & depth=2-4 & 0.02747 & 0.03798 & +0.01051 & A1 \\
5 & depth=2-8 & 0.02739 & 0.03450 & +0.00711 & A1 \\
6 & cx=0.80, mut=0.10 & 0.02760 & 0.03787 & +0.01027 & A1 \\
7 & cx=0.50, mut=0.30, hoist=0.10 & 0.02695 & 0.03278 & +0.00583 & A1 \\
\hline
\end{tabular}}
\caption{Comparison between Assignment 1 canonical GP and Assignment 2 SBGP mean test RMSE. Positive differences indicate that SBGP had higher error.}
\label{tab:a1-a2-comparison}
\end{table}

\subsection{Comparison with Assignment 1}

Compared with the canonical GP results from Assignment 1, the SBGP implementation did not improve average test performance. Assignment 1 produced lower mean test RMSE for every comparable configuration, as shown in Table~\ref{tab:a1-a2-comparison}. The closest result was the \texttt{pop=200} configuration, where SBGP had a mean test RMSE of 0.03089 compared with 0.02674 for canonical GP. This suggests that the larger population helped SBGP explore enough structures to reduce the performance gap, but it still did not outperform canonical GP on average.

\subsection{Runtimes}

\begin{table}[h]
\centering
\resizebox{\textwidth}{!}{
\begin{tabular}{llcccc}
\hline
\textbf{\#} & \textbf{Configuration} & \textbf{Average runtime (ms)} & \textbf{StdDev runtime (ms)} & \textbf{Fastest (ms)} & \textbf{Slowest (ms)} \\
\hline
1 & Baseline & 3908 & 2667 & 1712 & 10638 \\
2 & pop=50 & 2319 & 1633 & 1325 & 8066 \\
3 & pop=200 & 10209 & 10392 & 4450 & 47701 \\
4 & depth=2-4 & 2574 & 751 & 1188 & 4598 \\
5 & depth=2-8 & 5212 & 2127 & 3618 & 11049 \\
6 & cx=0.80, mut=0.10 & 3320 & 1120 & 1668 & 5483 \\
7 & cx=0.50, mut=0.30, hoist=0.10 & 3207 & 683 & 2346 & 4953 \\
\hline
\end{tabular}}
\caption{SBGP runtime results across 15 runs per configuration.}
\label{tab:runtime-results}
\end{table}

The runtime results in Table~\ref{tab:runtime-results} show the expected effect of population size. The \texttt{pop=50} configuration was fastest on average, with an average runtime of 2319 ms, while \texttt{pop=200} was slowest, with an average runtime of 10209 ms. The larger population also had the highest runtime variation because one run took 47701 ms. This is likely due to the cost of evaluating larger or more complex trees over the full training set. The depth setting also affected runtime: \texttt{depth=2-8} was slower than \texttt{depth=2-4}, which is expected because deeper trees can contain more nodes and therefore require more recursive evaluations.

The main finding is that SBGP was less reliable than canonical GP for this data set and parameter range. The structure-based global/local search can be beneficial when the selected global area is useful, but it can also cause the population to spend time exploiting a weak structural region. This appears to have happened in several runs, where SBGP converged to poorer expressions and increased the mean and standard deviation. The high standard deviations in Table~\ref{tab:sbgp-results}, especially for \texttt{pop=50}, \texttt{cx=0.80, mut=0.10}, and the baseline configuration, support this interpretation.

The best SBGP programs were still mostly based on lagged load variables such as \texttt{load\_d1}, \texttt{load\_d6}, and \texttt{load\_d7}. This is sensible because residential electricity demand has strong temporal structure, and the same time on previous days is highly informative. Therefore, SBGP did discover relevant predictors, but the structural restrictions did not consistently improve the search beyond canonical GP.

Overall, Assignment 2 did not outperform Assignment 1 in terms of average test RMSE. This does not mean that SBGP is generally worse than canonical GP; it means that for this implementation, data set, and parameter settings, the structure-based search was less stable. Further tuning of the SBGP-specific parameters, especially the global area depth $D_g$, similarity threshold $T_g$, and local search tolerance $W_g$, may improve reliability. A separate validation set could also be used to select the final model without relying only on training RMSE.

\end{document}
